A linear inside-outside algorithm for correcting sequencing errors in structured RNA

### 2013-01-09 Recomb/introduction_RECOMB.tex
recomb typos

For half a century, biological molecules have been studied as a proxy to understand evolution~\cite{Zuckerkandl1965}, and due
to their fundamental functions and remarkably conserved structures, rRNAs have always been a prime candidate for phylogenetic
studies~\cite{Olsen1986, Olsen1993}. In recent years, studies as the \emph{Human Microbiome Project}~\cite{Turnbaugh2007} benefited
of new technologies such as the NGS techniques to sequence as many new organisms as possible and extract an unprecedented flow of new information. 
Nonetheless, these high-throughput techniques typically have high error rates that make their applications to metagenomics (a.k.a. environmental
genomics) studies challenging. For instance, pyrosequencing as implemented by Roche's 454 produces may have a error rate raising up to 10\%.
Because there is no cloning step, resequencing to increase accuracy is not possible and it is therefore vital to disentangle noise from true sequence
diversity in this type of data \cite{Quince:2009uq}. Errors can be significantly reduced  when large multiple sequence alignments with close homologs
are available, but in studies of new or not well known organisms, such information is rather sparse. In particular, it is common to is not enough  similarity to 
differentiate between the sequencing errors and the natural polymorphisms that we want to observe, often inflating the diversity estimates~\cite{Kunin2010}.
A few techniques have been developed to remedy to this problem~\cite{Quinlan2008,Medvedev2011} but they do not take
into account all the available information. It is therefore essential to develop methods that can exploit any type of signal available to correct errors.

In this paper, we introduce \RNApyro, a novel algorithm to that enable us to calculate precisely mutational probabilities in RNA sequences with a
conserved consensus secondary structure. We show how our techniques can exploit the structural information embedded in physics-based energy models, 
covariance models and isostericity scales to identify and correct point-wise errors in RNA molecules with conserved secondary structure. In particular, we 
hypothesize that  conserved consensus secondary structures combined with sequence profiles  provide an information that allow us to identify and fix sequencing errors.

>>>>

For half a century, biological molecules have been studied as a proxy to understand evolution~\cite{Zuckerkandl1965}, and due
to their fundamental functions and remarkably conserved structures, rRNAs have always been a prime candidate for phylogenetic
studies~\cite{Olsen1986, Olsen1993}. In recent years, studies as the \emph{Human Microbiome Project}~\cite{Turnbaugh2007} benefited
of new technologies such as the NGS techniques to sequence as many new organisms as possible and extract an unprecedented flow of new information. 
Nonetheless, these high-throughput techniques typically have high error rates that make their applications to metagenomics (a.k.a. environmental
genomics) studies challenging. For instance, pyrosequencing as implemented by Roche's 454 produces may have an error rate raising up to 10\%.
Because there is no cloning step, resequencing to increase accuracy is not possible and it is therefore vital to disentangle noise from true sequence
diversity in this type of data \cite{Quince:2009uq}. Errors can be significantly reduced  when large multiple sequence alignments with close homologs
are available, but in studies of new or not well known organisms, such information is rather sparse. In particular, it is common that there is not enough  similarity to 
differentiate between the sequencing errors and the natural polymorphisms that we want to observe, often inflating the diversity estimates~\cite{Kunin2010}.
A few techniques have been developed to remedy to this problem~\cite{Quinlan2008,Medvedev2011} but they do not take
into account all the available information. It is therefore essential to develop methods that can exploit any type of signal available to correct errors.

In this paper, we introduce \RNApyro, a novel algorithm that enables us to calculate precisely mutational probabilities in RNA sequences with a
conserved consensus secondary structure. We show how our techniques can exploit the structural information embedded in physics-based energy models, 
covariance models and isostericity scales to identify and correct point-wise errors in RNA molecules with conserved secondary structure. In particular, we 
hypothesize that  conserved consensus secondary structures combined with sequence profiles  provide an information that allow us to identify and fix sequencing errors.

### 2013-01-09 Recomb/methods_RECOMB.tex
recomb typos

Let $s$ an RNA 
sequence and $m\geq 0$. $S$ is considered as  one derivation of the 
{SCFG} generating  
all  secondary structures of length  $|s|$. We are interested in  
the probability of a given position being a specific nucleotide,
over all sequences having at most $m$ mutations from $s$, under the SCFG derivation $S$ 
(formally 
$\mathbb{P}(s_i = x\mid s,\Omega, S,m)$). 
We define as a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990}, allowing us to obtain the
 desired probability,  the two  functions
$\Z{i,j}{m}{a,b}$ and $\Y{i,j}{m}{a,b}$. The former presented in 
Equations~\eqref{eq:Z_in} and~\eqref{eq:Z_rec}, our 
version of the \emph{inside} algorithm, computes the partition function between $i$ and $j$ included, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$) and containing 
$m$ mutations. The latter in Equations~\eqref{eq:Y_in} and~\eqref{eq:Y_rec},
 computes the \emph{outside}, in particular,  
the partition function considering only between $[0,i]\cup[j,n-1]$
knowing  that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$) and containing 
 $m$ mutations outside.

\subsubsection{Definitions}
Let be $B:=\left\{\text{A},\text{C},\text{G},\text{U}\right\}$, the set of nucleotides.
Given $s\in B^n$ an RNA sequence, let $s_i$ be the nucleotide at position $i$. Let $\Omega$ be a set of un-gapped RNA sequences of
length $n$, and $S$ a secondary structure without pseudoknots. 
Formally, if $(i,j)$ and $(k,l)$ are base pairs in $S$, there is no overlapping extremities
 $\{i,j\}\cap \{k,l\}=\varnothing$ and either the intersection is empty 
 ($[i,j]\cap[k,l]=\varnothing$) or one is included in the other ($[k,l]\subset[i,j]$ or 
 $[i,j]\subset[k,l])$. Let $R$ be the Boltzmann constant, $T$ the temperature in Kelvin and
  the function $\delta$ such that: 
 $\forall a,a' \in B, \delta_{a,a'}:=\left\{\begin{array}{ll}
															1 & \text{If } a\equiv a'\\
															0 & \text{Otherwise.}
														\end{array}\right.$

Finally, we denote by $E_{(i,j),ab \to ab'}^{\Omega,\beta}$ the local contribution of a base-pair $(i,j)$ to the pseudo-energy, such that
\begin{equation}
  E_{(i,j),ab \to a'b'}^{\Omega,\beta}  = \alpha \cdot\ES^\beta_{a b \to a' b'}+(1-\alpha)\cdot\EI^{\Omega}_{(i,j),a'b'}.
\end{equation}

@@

\subsection{Complexity}
Equations~\eqref{eq:Z_rec} and~\eqref{eq:Y_rec} can be computed using dynamic programming, i.e. $\mathcal{Z}^{*}_{*}$ or $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and interval lengths, memorizing the results such that they become available in constant time during future stages of the computation. Furthermore, energy terms $E(\cdot)$ can be accessed in constant time (after a precomputation in $\Theta(n\cdot|\Omega|)$ for the isostericity part). Computing any given term therefore requires $\Theta(m)$ operations.

In principle, this computation should be performed $\Theta(m\cdot n^2)$ times to investigate any triplet $(m,i,j)$.
However, a close inspection of the recurrences reveal that the computation can be safely restricted to a subset of all intervals.
For instance, in the inside computation, such intervals $[i,j]$ do not break any base-pair, and their next position $j+1$ is either past the end of the sequence, or is base-paired prior to $i$. Similar constraints hold for the outside computation. Such constraints drastically limit the combinatorics of required computations, dropping from $\Theta(n^2)$ to $\Theta(n)$ the number of terms that need to be computed and stored. Consequently the overall complexity of the algorithm is $\Theta(n\cdot(|\Omega|+m^2))$ aritmetic operations and $\Theta(n\cdot(|\Omega|+m))$ memory.

>>>>

Let $s$ be an RNA sequence, $S$ a reference structure, and $m\geq 0$ a desired number of mutations. 
We are interested in  the probability that a given position contains a specific nucleotide, over all sequences having at most $m$ mutations from $s$
%  under the SCFG derivation $S$ 
(formally $\mathbb{P}(s_i = x\mid s,\Omega, S,m)$). 
We define a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990}, allowing us to obtain the
 desired probability,  the two  functions
$\mathcal{Z}_*^*$ and $\mathcal{Y}_*^*$.

The former, defined in Equations~\eqref{eq:Z_in} and~\eqref{eq:Z_rec}, is analogous to the \emph{inside} algorithm.
It is the partition function, i.e. the sum of Boltzmann factors, over all sequences within $[i,j]$, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$), within 
$m$ mutations of $s$. The latter, defined by Equations~\eqref{eq:Y_in} and~\eqref{eq:Y_rec},
 computes the \emph{outside} algorithm, i.e.  
the partition function over sequences within $m$ mutations of $s$, restricted to two intervals $[0,i]\cup[j,n-1]$, and 
knowing  that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$). A suitable combination of these terms, given in Equation~\eqref{eq:combine}, gives the total weight, and in turn the probability, of seeing a specific base at a given position.

%Drawing a parallel with stochastic context-free grammars (SCFG), for which the inside-outside algorithm was introduced, the set of sequences can be seen as the language of words having length $n$, generated from a stochastic context-free grammar.
%These rules would limit the number of mutations (i.e. )
% $S$ can be considered as constraining weighing on the shape of parse trees the derivations in a classic 
% of the {SCFG} generating  all  secondary structures of length  $|s|$.

\subsubsection{Definitions}
Let $B:=\left\{\Ab,\Cb,\Gb,\Ub\right\}$ be the set of nucleotides.
Given $s\in B^n$ an RNA sequence, let $s_i$ be the nucleotide at position $i$. Let $\Omega$ be a set of un-gapped RNA sequences of
length $n$, and $S$ a secondary structure without pseudoknots. 
Formally, if $(i,j)$ and $(k,l)$ are base pairs in $S$, there is no overlapping extremities
 $\{i,j\}\cap \{k,l\}=\varnothing$ and either the intersection is empty 
 ($[i,j]\cap[k,l]=\varnothing$) or one is included in the other ($[k,l]\subset[i,j]$ or 
 $[i,j]\subset[k,l])$.

Let us then remind the Hamming distance function $\delta: B^*\times B^* \to \mathbb{N}^+$, which takes two sequences $s'$ and $s''$ as input, $|s'|=|s''|$, and returns the number of differing positions.
Finally, let us denote by $E_{(i,j),ab \to ab'}^{\Omega,\beta}$ the local contribution of a base-pair $(i,j)$ to the pseudo-energy, such that
\begin{equation}
  E_{(i,j),ab \to a'b'}^{\Omega,\beta}  = \alpha \cdot\ES^\beta_{a b \to a' b'}+(1-\alpha)\cdot\EI^{\Omega}_{(i,j),a'b'}.
\end{equation}

@@

\subsection{Complexity considerations}
Equations~\eqref{eq:Z_rec} and~\eqref{eq:Y_rec} can be computed using dynamic programming. Namely, the $\mathcal{Z}^{*}_{*}$ and $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and interval lengths, memorizing the results as they become available to ensure constant-time access during later stages of the computation. Furthermore, energy terms $E(\cdot)$ can be accessed in constant time thanks to a simple precomputation (not described)  of the isostericity contributions in $\Theta(n\cdot|\Omega|)$. Computing any given term therefore requires $\Theta(m)$ operations.

In principle, $\Theta(m\cdot n^2)$ terms, identified by $(m,i,j)$ triplets, should be computed.
However, a close inspection of the recurrences reveals that the computation can be safely restricted to a subset of intervals $(i,j)$.
For instance, the inside algorithm only requires computing intervals $[i,j]$ that do not break any base-pair, and whose next position $j+1$ is either past the end of the sequence, or is base-paired prior to $i$. Similar constraints hold for the outside computation, resulting in a drastic limitation of the combinatorics of required computations, dropping from $\Theta(n^2)$ to $\Theta(n)$ the number of terms that need to be computed and stored. Consequently the overall complexity of the algorithm is $\Theta(n\cdot(|\Omega|+m^2))$ arithmetic operations and $\Theta(n\cdot(|\Omega|+m))$ memory.
